---
title: "Interpretations of Anova"
---

### 1. ANOVA (Analysis of Variance)

ANOVA is a statistical method used to analyze the differences among group means in a sample. It tests whether there are any significant differences between the means of two or more groups.

#### Key Points:

-   **Purpose**: To determine if there are any significant differences between the means of two or more groups.

-   **Types**: One-way ANOVA (for one independent variable), Two-way ANOVA (for two independent variables).

-   **Assumptions**: Normally distributed data, homogeneity of variances (tested using Levene's test), independence of observations.

-   **R Packages**: `stats`, `car`, `afex`

```{r}
# Load required packages
options(repos = c(CRAN = "https://cloud.r-project.org"))
library(stats)  # Base R package for ANOVA
library(car)    # For ANOVA with Type III sums of squares

# One-way ANOVA
model <- aov(mpg ~ cyl, data = mtcars)
summary(model)

# Two-way ANOVA (Factorial design)
model <- aov(mpg ~ cyl + vs + cyl:vs, data = mtcars)
summary(model)
```

#### Notes:

-   **Factorial designs:** involve manipulating multiple independent variables (IVs) across different levels to observe their combined effects on dependent variables (DVs).

-   **RM ANOVA:** is used when measurements are taken from the same subjects over multiple time points or conditions.

### 2. ANCOVA (Analysis of Covariance)

ANCOVA extends ANOVA by including one or more continuous variables (covariates) in addition to the categorical independent variable(s). It adjusts group means based on these covariates.

#### Key Points:

-   **Purpose**: To compare group means while statistically controlling for the effects of one or more covariates.

-   **Assumptions**: Linearity between covariates and dependent variable, homogeneity of regression slopes.

-   **R Packages**: `car`, `lmtest`

```{r}
model <- lm(mpg ~ wt + factor(am), data = mtcars)
summary(model)
```

#### Notes:

-   **Covariate (CV)**: Acts to adjust group means based on pretest or unexplained exogenous variables. Must be continuous and have a linear relationship with the dependent variable (DV).

-   **Reasons to use ANCOVA**: Provides more power and adjusts for confounding variables.

-   **Reasons not to use ANCOVA**: Badly chosen variables may obscure real differences. Also, do not use ANCOVA if your covariate is related to group membership!!

### 3. MANOVA (Multivariate Analysis of Variance)

MANOVA is used when there are multiple dependent variables (DVs) and one or more independent variables (IVs). It tests the simultaneous effect of multiple IVs on multiple DVs.

#### Key Points:

-   **Purpose**: To test the joint effect of multiple independent variables on multiple dependent variables.

-   **Assumptions**: Multivariate normality, homogeneity of covariance matrices.

-   **R Packages**: `MANOVA`, `car`, `multcomp`

```{r}
data("iris")
model <- manova(cbind(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width) ~ Species, data = iris)
summary(model)
```

#### Notes:

-   **Reasons to use MANOVA**: Tests interaction of multiple DVs and IVs, protects against type 1 error.

-   **Reasons not to use MANOVA**: Techniques do not answer all questions; still need separate ANOVAs for significance testing.

#### Roy-Bargmann Stepdown:

Roy-Bargmann stepdown procedure prioritizes dependent variables (DVs) in a sequence of ANOVAs and ANCOVAs to understand which variables explain group differences most significantly.

-   The highest priority DV is tested in a univariate ANOVA.

-   The next most important DVs are tested in an ANCOVA with the high priority DV acting as a CV.

-   Sometimes the lower order DV still has an effect but not in the stepdown analysis, this tells us that yes, the DV explains some of the variance, but has no unique variability with the treatment after adjusting with the higher order DV and it is not needed.

``` r
# dv1 and dv2 are dependent variables of interest

# Univariate ANOVA for dv1
model1 <- aov(dv1 ~ group, data = mydata)

# ANCOVA for dv2 with dv1 as covariate
model2 <- lm(dv2 ~ dv1 + group, data = mydata)
```

-   **Notes**

    -   To find out which DV explains group differences in means, test them with separate ANOVAs.

    -   To find out which groups caused an effect, use post-hoc comparisons

### 4. Repeated Measures ANOVA

RM ANOVA is used when measurements are taken from the same subjects over multiple time points or conditions.

#### Key Points:

-   **When k = 2:** Use paired t-test or ANCOVA

-   **When k \> 2:** Use RM ANOVA or Profile Analysis

``` r
model <- aov(score ~ time + group + time:group + Error(subject/time), data = mydata)
```

#### Sphericity Violation and Corrections

-   Sphericity violation occurs when the variances of the differences between all possible pairs of within-subject conditions are not equal.

-   Violation leads to inflated type 1 error

-   Corrective measures such as Mauchly's test and Greenhouse-Geisser correction (especially for small n) can be applied.

``` r
# Sphericity test and correction
model <- aov(score ~ condition + Error(subject/condition), data = mydata)
summary(model)

# Check for sphericity violation
summary(aov(score ~ condition + Error(subject/condition), data = mydata))

# Greenhouse-Geisser correction
summary(aov(score ~ condition + Error(subject/condition), data = mydata, Greenhouse-Geisser = TRUE))
```
