# Project Description

*This is a research project I developed during my Master's Degree in Environmental Psychology. It includes a hierarchical model that explores the cross-level dynamics within communities in the Netherlands.*

*I am still motivated about executing this project again in a different setting because it addresses contemporary challenges in ethnically heterogeneous neighborhoods and provides an unique approach to understanding social cohesion. It assesses the individual-level, cognitive processes in neighborhood behavior through spatial measurement, which can offer alternative evidence for policymakers' views toward contextual level action (e.g., developing economically vital cities, demolition, and housing renovation) (Van Kempen & Bolt, 2009). By bridging the gap between individual-level processes and neighborhood dynamics, this research has the potential to inform policy interventions and community development initiatives.*

-   ***Outcome Variable***

    -   *`boundary`: combined value of perceived neighborhood scale and perceived trust in neighbors*

-   ***Demographics**:*

    -   *`age`: Normally distributed continuous variable.*

    -   *`education`: Dichotomous variable*

    -   *`income`: Ordinal variable from 1 to 5.*

    -   *`ethnicity`: Categorical variable sampled from four categories "A", "B", "C", and "D".*

    -   *`length_res`: Normally distributed continuous variable for length of residency.*

-   ***Physical Disorder**:*

    -   *`physical_disorder`: Ordinal variable from 1 to 5.*

-   ***Perceived Collective Efficacy**:*

    -   *`perceived_collective_efficacy`: Normally distributed continuous variable.*

-   ***Intergroup Leader**:*

    -   *`leadership`: Dichotomous variable.*

-   ***Personalized Features**:*

    -   *`names`, `art`, `playgrounds`: Dichotomous variables*

### Step 1: Load Mapping data

```{r}
# Load the libraries
library(sf)
library(ggplot2)
library(dplyr)
library(geojsonio)
library(sp)

# Read the GeoJSON file
map_data <- geojson_read("map_data.geojson", what = "sp")

# Convert to sf object for easier handling
map_sf <- st_as_sf(map_data)

# Calculate the area of each drawn boundary
map_sf <- map_sf %>%
  mutate(area = st_area(geometry))

# Convert area to numeric
map_sf$area <- as.numeric(map_sf$area)

# Plot a sample of the boundaries
ggplot(data = map_sf) +
  geom_sf() +
  theme_minimal() +
  labs(title = "Neighborhood Boundaries")
```

### Step 2: Load Survey Data

```{r}
# Load necessary libraries
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("lme4")
install.packages("car")
install.packages("ggplot2")
library(lme4)
library(car)
library(ggplot2)

# Load the dataset
data <- read.csv("multilevel_model_data.csv")

# Merge datasets
data$id <- 1:nrow(data)
data <- data %>%
  left_join(map_sf %>% select(id, area), by = "id")
```

### Step 3: Exploratory Data Analysis

*Before I compare the models, let's take a look at the full model and understand out data a bit better.*

```{r}
# Fit the multilevel model
multilevel_model <- lmer(boundary ~ age + education + income + ethnicity + length_res + physical_disorder + perceived_collective_efficacy + facilities + names + art + playgrounds + leadership + (1 | neighborhood_id), data = data)

# Print summary of the model
summary(multilevel_model)

# Check normality of residuals
plot(multilevel_model)

# Check homoscedasticity
plot(residuals(multilevel_model) ~ fitted(multilevel_model))

# Check linearity (if applicable)
# Example for a continuous predictor age
plot(data$age, residuals(multilevel_model))

# Extract fixed effects estimates
fixed_effects <- fixef(multilevel_model)

# Interpretation of fixed effects
print(fixed_effects)
```

### Step 4: Comparing Alternative Models:

```{r}
# Model without Low-estimate predictors:
alt_model1 <- lmer(boundary ~ education + income + ethnicity + perceived_collective_efficacy + facilities + art + leadership + (1 | neighborhood_id), data = data)

# Individual-focused Model:
alt_model2 <- lmer(boundary ~ age + education + income + perceived_collective_efficacy + (1 | neighborhood_id), data = data)

# Interaction Effects:
alt_model3 <- lmer(boundary ~ age * perceived_collective_efficacy + education + income + length_res + physical_disorder + facilities + names + art + playgrounds + leadership + (1 | neighborhood_id), data = data)

# Random Slopes Model:
alt_model4 <- lmer(boundary ~ age + education + income + perceived_collective_efficacy + (age + education + income + perceived_collective_efficacy | neighborhood_id), data = data)

# Cross-Level Interactions:
alt_model5 <- lmer(boundary ~ age + education + income + perceived_collective_efficacy + facilities + names + art + playgrounds + leadership + (age + education + income + perceived_collective_efficacy | neighborhood_id), data = data)

# Non-linear Effects:
data$age_squared <- data$age^2
alt_model6 <- lmer(boundary ~ age + age_squared + education + income + perceived_collective_efficacy + (1 | neighborhood_id), data = data)

# Nested Models:
alt_model7 <- lmer(boundary ~ age + education + income + perceived_collective_efficacy + (1 | neighborhood_id) + (1 + age | neighborhood_id), data = data)

# Compare models using AIC
AIC_values <- AIC(multilevel_model, alt_model1, alt_model2, alt_model3, alt_model4, alt_model5, alt_model6, alt_model7)
BIC_values <- BIC(multilevel_model, alt_model1, alt_model2, alt_model3, alt_model4, alt_model5, alt_model6, alt_model7)
print(AIC_values)
print(BIC_values)
```

*After removing variables (`age, length_res, names, playgrounds, physical_disorder)` , my first adjusted model has one of the lower AIC/BIC, while also aligning with my theoretical framework and hypotheses.*

### Step 5: Post-estimation Analyses:

```{r}
influencePlot(alt_model1, id.n = 5)

# Calculate variance inflation factors (VIF)
vif(alt_model1)
```

### Step 6: Reporting:

```{r}
# Extract coefficients
model_coefs <- coef(summary(alt_model1))
model_coefs
```

### References

*Bektaş, Y., & Taşan-Kok, T. (2020). Love thy neighbor? Remnants of the social-mix policy in the Kolenkit neighborhood, Amsterdam. Journal of Housing and the Built Environment, 35, 743-761.*

*Campbell, E., Henly, J. R., Elliott, D. S., & Irwin, K. (2009). Subjective constructions of neighborhood boundaries: lessons from a qualitative study of four neighborhoods. Journal of Urban Affairs, 31(4), 461-490.*

*Chan, J., To, H. P., & Chan, E. (2006). Reconsidering social cohesion: Developing a definition and analytical framework for empirical research. Social Indicators Research, 75, 273-302.*

*Coulton, C. J., Jennings, M. Z., & Chan, T. (2013). How big is my neighborhood? Individual and contextual effects on perceptions of neighborhood scale. American Journal of Community Psychology, 51(1-2), 140-150.*

*Easterbrook, M. J., & Vignoles, V. L. (2015). When friendship formation goes down the toilet: Design features of shared accommodation influence interpersonal bonds and well‐being. British Journal of Social Psychology, 54(1), 125-139.*

*Jordan, B. (1996). A theory of poverty & social exclusion.*

*Kleinhans, R. (2004). Social implications of housing diversification in urban renewal: A review of recent literature. Journal of Housing and the Built Environment, 19, 367-390.*

*Kloos, B., Hill, J., Thomas, E., Wandersman, A., Elias, M. J., & Dalton, J. H. (2012). Community psychology. Belmont, CA: Cengage Learning.*

*Lowe, S. S. (2000). Creating community: Art for community development. Journal of Contemporary Ethnography, 29(3), 357-386.*

*Lynch, S. M., & Bartlett, B. (2019). Bayesian statistics in sociology: Past, present, and future. Annual Review of Sociology, 45, 47-68.*

*Peng, J., Strijker, D., & Wu, Q. (2020). Place identity: How far have we come in exploring its meanings?. Frontiers in Psychology, 11, 503569.*

*Pretty, G. H., Chipuer, H. M., & Bramston, P. (2003). Sense of place amongst adolescents and adults in two rural Australian towns: The discriminating features of place attachment, sense of community and place dependence in relation to place identity. Journal of environmental psychology, 23(3), 273-287.*

*Putnam, R. D. (2000). Bowling alone: The collapse and revival of American community. Simon and Schuster.*

*Putnam, R. D. (2007). E pluribus unum: Diversity and community in the twenty‐first century the 2006 Johan Skytte Prize Lecture. Scandinavian Political Studies, 30(2), 137-174.*

*Stevenson, C., Easterbrook, M., Harkin, L., McNamara, N., Kellezi, B., & Shuttleworth, I. (2019). Neighborhood identity helps residents cope with residential diversification: Contact in increasingly mixed neighborhoods of Northern Ireland. Political Psychology, 40(2), 277-295.*

*Stolle, D., & Harell, A. (2013). Social capital and ethno-racial diversity: Learning to trust in an immigrant society. Political Studies, 61(1), 42-66.*

*Taylor, R. B., Gottfredson, S. D., & Brower, S. (1984). Neighborhood naming as an index of attachment to place. Population and Environment, 7, 103-125.*

*Turner, R. N., Hewstone, M., Voci, A., & Vonofakou, C. (2008). A test of the extended intergroup contact hypothesis: The mediating role of intergroup anxiety, perceived ingroup and outgroup norms, and inclusion of the outgroup in the self. Journal of Personality and Social Psychology, 95(4), 843.*

*Van Kempen, R., & Bolt, G. (2009). Social cohesion, social mix, and urban policies in the Netherlands. Journal of Housing and the Built Environment, 24, 457-475.*

*Völker, B., & Flap, H. (2007). Sixteen million neighbors: A multilevel study of the role of neighbors in the personal networks of the Dutch. Urban Affairs Review, 43(2), 256-284.*
