# Project Description

*This is a research project I developed during my Master's Degree in Environmental Psychology. It includes a hierarchical model that explores the cross-level dynamics within communities in the Netherlands.*

*I am still motivated about executing this project again in a different setting because it addresses contemporary challenges in ethnically heterogeneous neighborhoods and provides an unique approach to understanding social cohesion. It assesses the individual-level, cognitive processes in neighborhood behavior through spatial measurement, which can offer alternative evidence for policymakers' views toward contextual level action (e.g., developing economically vital cities, demolition, and housing renovation) (Van Kempen & Bolt, 2009). By bridging the gap between individual-level processes and neighborhood dynamics, this research has the potential to inform policy interventions and community development initiatives.*

-   ***Outcome Variable***

    -   *`boundary`: combined value of perceived neighborhood scale and perceived trust in neighbors*

-   ***Demographics**:*

    -   *`age`: Normally distributed continuous variable.*

    -   *`education`: Dichotomous variable*

    -   *`income`: Ordinal variable from 1 to 5.*

    -   *`ethnicity`: Categorical variable sampled from four categories "A", "B", "C", and "D".*

    -   *`length_res`: Normally distributed continuous variable for length of residency.*

-   ***Physical Disorder**:*

    -   *`physical_disorder`: Ordinal variable from 1 to 5.*

-   ***Perceived Collective Efficacy**:*

    -   *`perceived_collective_efficacy`: Normally distributed continuous variable.*

-   ***Intergroup Leader**:*

    -   *`leadership`: Dichotomous variable.*

-   ***Personalized Features**:*

    -   *`names`, `art`, `playgrounds`: Dichotomous variables*

### 1. Load Packages and Data

```{r}
# Load libraries
options(repos = c(CRAN = "https://cloud.r-project.org"))
library(sf)
library(ggplot2)
library(dplyr)
library(geojsonio)
library(sp)
library(lme4)
library(car)
library(ggplot2)

# Read the GeoJSON mapping file
map_data <- geojson_read("map_data.geojson", what = "sp")

map_sf <- st_as_sf(map_data)

# Calculate the area of each drawn boundary
map_sf <- map_sf %>%
  mutate(area = st_area(geometry))
map_sf$area <- as.numeric(map_sf$area)
```

```{r}
# Load the survey dataset
data <- read.csv("multilevel_model_data.csv")

# Merge datasets
data$id <- 1:nrow(data)
data <- data %>%
  left_join(map_sf %>% select(id, area), by = "id")
```

### 2. Exploratory Data Analysis

*Before I compare the models, let's take a look at the full model and understand our data a bit better.*

```{r}
# Fit the multilevel model
multilevel_model <- lmer(boundary ~ age + education + income + ethnicity + length_res + physical_disorder + perceived_collective_efficacy + facilities + names + art + playgrounds + leadership + (1 | neighborhood_id), data = data)
summary(multilevel_model)

# Normality of residuals
plot(multilevel_model)

# Homoscedasticity
plot(residuals(multilevel_model) ~ fitted(multilevel_model))

# Fixed effects estimates
fixed_effects <- fixef(multilevel_model)
print(fixed_effects)
```

### 3. Model Selection

```{r}
# Model without Low-estimate predictors:
alt_model1 <- lmer(boundary ~ education + income + ethnicity + perceived_collective_efficacy + facilities + art + leadership + (1 | neighborhood_id), data = data)

# Individual-focused Model:
alt_model2 <- lmer(boundary ~ age + education + income + perceived_collective_efficacy + (1 | neighborhood_id), data = data)

# Interaction Effects:
alt_model3 <- lmer(boundary ~ age * perceived_collective_efficacy + education + income + length_res + physical_disorder + facilities + names + art + playgrounds + leadership + (1 | neighborhood_id), data = data)

# Random Slopes Model:
alt_model4 <- lmer(boundary ~ age + education + income + perceived_collective_efficacy + (age + education + income + perceived_collective_efficacy | neighborhood_id), data = data)

# Cross-Level Interactions:
alt_model5 <- lmer(boundary ~ age + education + income + perceived_collective_efficacy + facilities + names + art + playgrounds + leadership + (age + education + income + perceived_collective_efficacy | neighborhood_id), data = data)

# Non-linear Effects:
data$age_squared <- data$age^2
alt_model6 <- lmer(boundary ~ age + age_squared + education + income + perceived_collective_efficacy + (1 | neighborhood_id), data = data)

# Nested Models:
alt_model7 <- lmer(boundary ~ age + education + income + perceived_collective_efficacy + (1 | neighborhood_id) + (1 + age | neighborhood_id), data = data)

# Compare models using AIC
AIC_values <- AIC(multilevel_model, alt_model1, alt_model2, alt_model3, alt_model4, alt_model5, alt_model6, alt_model7)
BIC_values <- BIC(multilevel_model, alt_model1, alt_model2, alt_model3, alt_model4, alt_model5, alt_model6, alt_model7)
print(AIC_values)
print(BIC_values)
```

*After removing variables (`age, length_res, names, playgrounds, physical_disorder)` , my first adjusted model has one of the lower AIC/BIC, while also aligning with my theoretical framework and hypotheses.*

### 4. Results

```{r}
# Plot Cooks D estimates with influence plot
influencePlot(alt_model1, id.n = 5)

# Calculate variance inflation factors
vif(alt_model1)

# Extract coefficients
model_coefs <- coef(summary(alt_model1))
model_coefs
```
