# Project Description

*This is a research project I developed during my Master's Degree in Environmental Psychology. It includes an introduction and a hierarchical model that explores the cross-level dynamics within communities in the Netherlands.*

*I am still motivated about executing this project in a practical setting because it addresses contemporary challenges in ethnically heterogeneous neighborhoods and provides an unique approach to understanding social cohesion. It assesses the individual-level, cognitive processes in neighborhood behavior through spatial measurement, which can offer alternative evidence for policymakers' views toward contextual level action (e.g., developing economically vital cities, demolition, and housing renovation) (Van Kempen & Bolt, 2009). By bridging the gap between individual-level processes and neighborhood dynamics, this research has the potential to inform policy interventions and community development initiatives.*

***Title: How Individual Level Processes Undermine Social Cohesion in Ethnically Heterogeneous Neighborhoods***

***Abstract***

*Current research on ethnically heterogeneous neighborhoods lacks individual-level variables, limiting evaluations of social cohesion. Building upon Bektaş and Taşan-Kok (2020), this proposal integrates psychological constructs like place identity and intergroup anxiety to enhance the current framework. By utilizing cognitive mapping techniques, I aim to provide a more comprehensive understanding of social cohesion at the intersection of individual and neighborhood factors. Through hierarchical modeling, this study seeks to assess cross-level interactions and variability across ethnically heterogeneous neighborhoods in the Netherlands.*\

***Introduction***

*Modern individualized lifestyles have forgotten the special nature of an intimate society (Kloos et al., 2012). Between 1970 and 2000, close-knit neighborhoods were more common, and they enriched our social lives with friendly routine discourse and spontaneous activities (Putnam, 2000). Residents were generally happier, safer, and healthier (Jordan, 1996). More importantly, they had place identities; a cognitive overlap with the self and the community (Turner et al., 2008). Strong place identities led people to treat neighbors like the self by embodying their problems and successes, which helped create a sense of social cohesion (Stevenson et al., 2019; Turner et al., 2008). However, in recent decades, there has been a noticeable decline in neighborhood engagement, contributing to increased social isolation among individuals (Kloos et al., 2012; Putnam, 2000). This trend is particularly pronounced in the Netherlands, where personal networks consist of fewer neighbors (Völker & Flap, 2007).*

*Social cohesion, often influenced by residents' ethnic similarities (Van Kempen & Bolt, 2009), facilitates the development of strong place identities (Putnam, 2000; Jordan, 1996). But in ethnically heterogeneous neighborhoods, individual differences create strenuous interactions that can lead to social dissolution (Stevenson et al., 2019), which is defined as the opposite of social cohesion (Chan et al., 2006). Social dissolution was found in Northern Ireland, where residents in religiously desegregated neighborhoods had trouble defining their place identities (Stevenson et al., 2019). With no sense of belonging, they struggle to form strong relationships with each other. More recently, Bektaş and Taşan-Kok (2020) found similar trends in the Kolenkit neighborhood located in Amsterdam, and residents placed most of the blame on the cultural differences. With increased migration on the rise in the Netherlands (Van Kempen & Bolt, 2009), the inevitable reality of ethnically heterogeneous neighborhoods should reveal the criticality of place identity research.*

*Policy interventions often overlook the individual-level processes (e.g., place identity), relying on social mixing policies that fail to address underlying issues (Bektaş & Taşan-Kok, 2020; Van Kempen & Bolt, 2009). In 2007, the Minister of Housing, Districts, and Integration prompted the integration of different social and ethnic groups into neighborhoods with confidence that designing innovative public spaces or identifying the perfect distribution of ethnicities per district would promote social cohesion (Van Kempen & Bolt, 2009). However, direct intergroup contact does not always promote social cohesion, it can sometimes damage relationships even more by creating intergroup anxiety (Turner et al., 2008). Concerns raised by researchers in the Netherlands highlight the need for more nuanced approaches (Kleinhans 2004; Stolle & Harell, 2013; Van Kempen & Bolt, 2009; Völker & Flap, 2007). For instance, Bektaş and Taşan-Kok (2020) found that the social mixing policies only led to superficial interactions that had little impact on social cohesion. Similarly, Stevenson et al. (2019) found that close proximity was insufficient to foster positive intergroup relationships, emphasizing the importance of developing place identity.*

*My research proposal aims to complement Bektaş and Taşan-Kok (2020) and Stevenson et al. (2019) by studying the individual-level processes that undermine social cohesion in ethnically heterogeneous neighborhoods in the Netherlands. The problem at hand is the presence of unmeasured individual-level variables contributing to the social dissolution of these neighborhoods, thereby rendering contemporary evaluations of social cohesion inaccurate. This research seeks to address the following questions: (1) What reliable measures of social cohesion can capture the unexplained variability at the individual level? (2) What is the current state of residents' place identity in ethnically heterogeneous neighborhoods? (3) How can multilevel evidence about individual and contextual variables shed light on social cohesion and intergroup anxiety? (4) What impact does neighborhood leadership have on reducing intergroup anxiety? (5) How do personalized neighborhood features influence place identity?\
*

***Theoretical Approach***

*The principles of intergroup contact offer insights into the challenges faced by ethnically heterogeneous neighborhoods in maintaining social cohesion. Introducing an outgroup minority to an ingroup native community can trigger intergroup anxiety---a response rooted in negative expectations of rejection or discrimination during interactions (Turner et al., 2008). This anxiety can result in social isolation and narrowed attention (Van Kempen & Bolt, 2009). While direct interactions do not invariably lead to positive attitudes, intergroup anxiety can be mitigated by factors like pre-established friendships. Pre-established friendships foster inclusivity and reduce uncertainty for outgroup members, facilitating positive contact (Turner et al., 2008). In the context of this proposal, the presence of inclusive members or leaders within the ingroup community, and showing interest in the outgroup culture, may mitigate intergroup anxiety.*

*Strong place identities emerge when residents feel a sense of belonging, aided by emotional ties to neighborhood features (Peng et al., 2020; Stevenson et al., 2019). Measuring place identity has multiple avenues, and some suggest that one or two-item scales are adequate (Peng et al., 2020). In community-based research, small-item scales may be a preferable method that would lead to more responses (Pretty et al., 2003).*

*Measuring social cohesion poses challenges, as direct inquiries about neighbors may only capture part of the picture (Bektaş & Taşan-Kok, 2020). Cognitive mapping offers a promising avenue to understand residents' spatial perceptions (Campbell et al., 2009). Coulton et al. (2013) applied this strategy when studying the effect a neighborhood has on individuals in America, where residents drew boundaries around their perceived neighborhood on a blank map of their district, reflecting disagreements between them. In regards to intergroup anxiety, strong neighborhood ties can be assessed by studying mutual trust (Bektaş & Taşan-Kok, 2020; Stolle & Harell, 2013; Turner et al., 2008). For example, measuring residents' perceived boundary of trust in regards to theft may reveal previously unexplained variability in intergroup anxiety.\
*

***Individual Predictors***

*At the individual level, disparities in education, length of residency in the neighborhood, and income have been linked to social dissolution (Coulton et al., 2013; Stevenson et al., 2019; Van Kempen & Bolt, 2009), potentially affecting perceived boundaries. Additionally, Coulton et al. (2013) advocate for assessing individual perceptions of collective efficacy---a measure of confidence in neighborhood action (Sampson et al., 1997). Collective efficacy encompasses the belief that others will contribute, influencing one's engagement in community activities (Kloos et al., 2012).*

***Contextual Predictors***

*At the contextual level, neighborhood characteristics, which help develop place identity, are also associated with positive intergroup attitudes (Stevenson et al., 2019). People generally prefer to live in orderly areas that contain operational facilities (Coulton et al., 2013). Personalizing neighborhood features such as community names (Stevenson et al., 2019; Taylor et al., 1984), local art (Lowe, 2000), or playgrounds (Kloos et al., 2012) can help people develop place identities toward their neighborhood. As mentioned, intergroup friendships and/or leadership can impact an outgroup member's sense of belonging. Inclusive leadership, as a contextual level predictor, may shape intergroup attitudes. However, it is critical to note that individual-level processes can still influence these features. For example, public spaces can be effective because they provide an area for close interactions (Easterbrook et al., 2015), but this can be less effective in intergroup settings (Völker & Flap, 2007). Exploring cross-level interactions would help clear up some of the contradictions in the literature.*

***Procedure***

*Residents will be asked to draw their perceived neighborhood scale (social cohesion) and perceived trust in neighbors (intergroup anxiety) on a blank map of the district. This can be completed through a door-to-door survey or an online questionnaire. More specifically, they will draw a boundary on a map that encompasses their neighborhood boundaries and the perceived space in which they would feel safe leaving their personal belongings unattended for 24 hours, excluding individuals from outside their neighborhood who could potentially be involved in theft.*

*GIS software can be used to estimate residuals when analyzing mapping data, as suggested by Coulton et al. (2013). Alternatively, R programming, supported by packages such as "ggmap," "choroplethr," and "choroplethrMaps," can be utilized for this purpose. Given my extensive familiarity with R software, it will be the chosen tool for analysis.*

-   ***Demographics**:*

    -   *`age`: Normally distributed continuous variable.*

    -   *`education`: Dichotomous variable generated using `rbinom` with a probability of 0.6 for university education.*

    -   *`income`: Ordinal variable sampled from 1 to 5.*

    -   *`ethnicity`: Categorical variable sampled from four categories "A", "B", "C", and "D".*

    -   *`length_res`: Normally distributed continuous variable for length of residency.*

-   ***Physical Disorder**:*

    -   *`physical_disorder`: Ordinal variable sampled from 1 to 5.*

-   ***Perceived Collective Efficacy**:*

    -   *`perceived_collective_efficacy`: Normally distributed continuous variable.*

-   ***Intergroup Leader**:*

    -   *`leadership`: Dichotomous variable generated using `rbinom`.*

-   ***Personalized Features**:*

    -   *`names`, `art`, `playgrounds`: Dichotomous variables generated using `rbinom`.*

### 

### Step 1: Install and load necessary packages

```{r}
# Load necessary libraries
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("lme4")
install.packages("car")
install.packages("ggplot2")
library(lme4)
library(car)
library(ggplot2)

# Load the dataset
data <- read.csv("multilevel_model_data.csv")
```

### Step 2: Exploratory Data Analysis

*Before I compare the models, let's take a look at the full model and understand out data a bit better.*

```{r}
# Fit the multilevel model
multilevel_model <- lmer(boundary ~ age + education + income + ethnicity + length_res + physical_disorder + perceived_collective_efficacy + facilities + names + art + playgrounds + leadership + (1 | neighborhood_id), data = data)

# Print summary of the model
summary(multilevel_model)

# Check normality of residuals
plot(multilevel_model)

# Check homoscedasticity
plot(residuals(multilevel_model) ~ fitted(multilevel_model))

# Check linearity (if applicable)
# Example for a continuous predictor age
plot(data$age, residuals(multilevel_model))

# Extract fixed effects estimates
fixed_effects <- fixef(multilevel_model)

# Interpretation of fixed effects
print(fixed_effects)
```

### Step 3: Comparing Alternative Models:

```{r}
# Model without Low-estimate predictors:
alt_model1 <- lmer(boundary ~ education + income + ethnicity + perceived_collective_efficacy + facilities + art + leadership + (1 | neighborhood_id), data = data)

# Individual-focused Model:
alt_model2 <- lmer(boundary ~ age + education + income + perceived_collective_efficacy + (1 | neighborhood_id), data = data)

# Interaction Effects:
alt_model3 <- lmer(boundary ~ age * perceived_collective_efficacy + education + income + length_res + physical_disorder + facilities + names + art + playgrounds + leadership + (1 | neighborhood_id), data = data)

# Random Slopes Model:
alt_model4 <- lmer(boundary ~ age + education + income + perceived_collective_efficacy + (age + education + income + perceived_collective_efficacy | neighborhood_id), data = data)

# Cross-Level Interactions:
alt_model5 <- lmer(boundary ~ age + education + income + perceived_collective_efficacy + facilities + names + art + playgrounds + leadership + (age + education + income + perceived_collective_efficacy | neighborhood_id), data = data)

# Non-linear Effects:
data$age_squared <- data$age^2
alt_model6 <- lmer(boundary ~ age + age_squared + education + income + perceived_collective_efficacy + (1 | neighborhood_id), data = data)

# Nested Models:
alt_model7 <- lmer(boundary ~ age + education + income + perceived_collective_efficacy + (1 | neighborhood_id) + (1 + age | neighborhood_id), data = data)

# Compare models using AIC
AIC_values <- AIC(multilevel_model, alt_model1, alt_model2, alt_model3, alt_model4, alt_model5, alt_model6, alt_model7)
BIC_values <- BIC(multilevel_model, alt_model1, alt_model2, alt_model3, alt_model4, alt_model5, alt_model6, alt_model7)
print(AIC_values)
print(BIC_values)
```

*After removing variables (`age, length_res, names, playgrounds, physical_disorder)` , my first adjusted model has one of the lower AIC/BIC, while also aligning with my theoretical framework and hypotheses.*

### Step 4: Post-estimation Analyses:

```{r}
influencePlot(alt_model1, id.n = 5)

# Calculate variance inflation factors (VIF)
vif(alt_model1)
```

### Step 5: Reporting:

```{r}
# Extract coefficients
model_coefs <- coef(summary(alt_model1))
model_coefs
```

-   *`art` and `leadership` show to have a strong influence on community boundaries and hence place identity.*
-   It is difficult to make any strong claims due to the lack of reliability in the data, but further research could lead to groundbreaking information as to how interracial communities can form social cohesion.

### References

*Bektaş, Y., & Taşan-Kok, T. (2020). Love thy neighbor? Remnants of the social-mix policy in the Kolenkit neighborhood, Amsterdam. Journal of Housing and the Built Environment, 35, 743-761.*

*Campbell, E., Henly, J. R., Elliott, D. S., & Irwin, K. (2009). Subjective constructions of neighborhood boundaries: lessons from a qualitative study of four neighborhoods. Journal of Urban Affairs, 31(4), 461-490.*

*Chan, J., To, H. P., & Chan, E. (2006). Reconsidering social cohesion: Developing a definition and analytical framework for empirical research. Social Indicators Research, 75, 273-302.*

*Coulton, C. J., Jennings, M. Z., & Chan, T. (2013). How big is my neighborhood? Individual and contextual effects on perceptions of neighborhood scale. American Journal of Community Psychology, 51(1-2), 140-150.*

*Easterbrook, M. J., & Vignoles, V. L. (2015). When friendship formation goes down the toilet: Design features of shared accommodation influence interpersonal bonds and well‐being. British Journal of Social Psychology, 54(1), 125-139.*

*Jordan, B. (1996). A theory of poverty & social exclusion.*

*Kleinhans, R. (2004). Social implications of housing diversification in urban renewal: A review of recent literature. Journal of Housing and the Built Environment, 19, 367-390.*

*Kloos, B., Hill, J., Thomas, E., Wandersman, A., Elias, M. J., & Dalton, J. H. (2012). Community psychology. Belmont, CA: Cengage Learning.*

*Lowe, S. S. (2000). Creating community: Art for community development. Journal of Contemporary Ethnography, 29(3), 357-386.*

*Lynch, S. M., & Bartlett, B. (2019). Bayesian statistics in sociology: Past, present, and future. Annual Review of Sociology, 45, 47-68.*

*Peng, J., Strijker, D., & Wu, Q. (2020). Place identity: How far have we come in exploring its meanings?. Frontiers in Psychology, 11, 503569.*

*Pretty, G. H., Chipuer, H. M., & Bramston, P. (2003). Sense of place amongst adolescents and adults in two rural Australian towns: The discriminating features of place attachment, sense of community and place dependence in relation to place identity. Journal of environmental psychology, 23(3), 273-287.*

*Putnam, R. D. (2000). Bowling alone: The collapse and revival of American community. Simon and Schuster.*

*Putnam, R. D. (2007). E pluribus unum: Diversity and community in the twenty‐first century the 2006 Johan Skytte Prize Lecture. Scandinavian Political Studies, 30(2), 137-174.*

*Stevenson, C., Easterbrook, M., Harkin, L., McNamara, N., Kellezi, B., & Shuttleworth, I. (2019). Neighborhood identity helps residents cope with residential diversification: Contact in increasingly mixed neighborhoods of Northern Ireland. Political Psychology, 40(2), 277-295.*

*Stolle, D., & Harell, A. (2013). Social capital and ethno-racial diversity: Learning to trust in an immigrant society. Political Studies, 61(1), 42-66.*

*Taylor, R. B., Gottfredson, S. D., & Brower, S. (1984). Neighborhood naming as an index of attachment to place. Population and Environment, 7, 103-125.*

*Turner, R. N., Hewstone, M., Voci, A., & Vonofakou, C. (2008). A test of the extended intergroup contact hypothesis: The mediating role of intergroup anxiety, perceived ingroup and outgroup norms, and inclusion of the outgroup in the self. Journal of Personality and Social Psychology, 95(4), 843.*

*Van Kempen, R., & Bolt, G. (2009). Social cohesion, social mix, and urban policies in the Netherlands. Journal of Housing and the Built Environment, 24, 457-475.*

*Völker, B., & Flap, H. (2007). Sixteen million neighbors: A multilevel study of the role of neighbors in the personal networks of the Dutch. Urban Affairs Review, 43(2), 256-284.*
